### 2017-06-06 sheets/talk-6/sheet-6.tex

\begin{exercise}[Coequalizer of sets]
  In this exercise we will first define \emph{reflexive symmetric transitive closures} and will define the coequalizer of $\Set$.
  \begin{itemize}
    \item[(a)]First we will show that the reflexive, symmetric and transitive closure of a relation $R$ is the \emph{least} equivalence relation containing $R$.
      Let $R$ be a relation over a set $X$ and $S$ be the reflexive, symmetric and transitive closure of $R$. Show following lemmas:
      \begin{itemize}
        \item[(i)]Show $\forall n\in\nat: R_n$ and $S$ are reflexive and symmetric.
        \item[(ii)]Show $\forall n\in\nat: R \subseteq R_n$ and $R \subseteq S$.
        \item[(iii)]Show that $R_n$ is \emph{monotone}, i.\,e. $\forall n,m\in\nat: n\le m \Rightarrow R_n \subseteq R_m$.
        \item[(iv)]Now show that $S$ is transitive.
          \begin{hint}
            Let $m\in\nat$. Show $\forall x_1, x_2, x_3: (x_1, x_2) \in S \Rightarrow (x_2, x_3) \in R_m \Rightarrow (x_1, x_3) \in S$ via $\nat$-induction.
          \end{hint}
        \item[(v)]Now show that $R$ is the \emph{least} equivalence relation containing $R$, i.\,e.
          $$\forall R': \text{$R'$ is a equivalence relation} \land R \subseteq R' \Rightarrow S \subseteq R'.$$
      \end{itemize}
    \item[(b)]Now we will define the coequalizer of $\Set$.
      Let $X, Y \in \Set, s, t \from X \to Y$.
      We make the following definitions:
      \begin{alignat*}{2}
        R    &:=& \set{(s(a), t(a)) | a\in Y} \\
        \sim &:=& \text{reflexive, symmetric and transitive closure of $R$} \\
        C    &:=& Y / \sim
      \end{alignat*}
      Note that $[y] := \set{ y' \in Y | (y, y') \in \sim}$ is the equivalence class of $y$ and $Y / \sim := \set{ [y] | y \in Y}$ is the set of all equivalence classes of $\sim$.
      Also note that no equivalence class can be empty and for $[y]$ we call $y$ a representant of this class.
      We define $i \from Y \to E$ with $i(y) := [y]$.
      \begin{itemize}
        \item[(i)]Show that the following diagramm commutes:
          $\xymatrix{
            X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
          }$
        \item[(ii)]Let
          $\xymatrix{
            X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
          }$
          be a cummutative diagramm. Show that $f$ respects equivalence, i.\,e. $\forall (y_1, y_2) \in \sim: f(y_1) = f(y_2)$.
          \begin{hint}
            ``Destruct'' $(y_1, y_2) \in \sim$ and do a $\nat$-induction.
          \end{hint}
        \item[(iii)]Now define $\bar{f} \from C \to A$, such that the following diagramm commutes and show that there can only be one:
          \[ \xymatrix{
            & & A  \\
            X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[ru]^f \ar[rd]_i \\
            & & C \ar@{.>}[uu]|{\bar{f}}
          } \]
        \item[(iv)]Show that the fork
          $\xymatrix{
            X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
          }$
          is the coequalizer of
          $\xymatrix{
            X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
          }$.
      \end{itemize}
  \end{itemize}
\end{exercise}

>>>>

\newcommand{\choice}[1]{\{ #1 \}}
\newcommand{\class} [1]{[ #1 ]}

\begin{exercise}[Coequalizer of sets]
  In this exercise we will will define the coequalizer of $\Set$ with the help of reflexive, symmetric and transitive closures.
  Let $X, Y \in \Set, s, t \from X \to Y$.
  We make the following definitions:
  \begin{alignat*}{2}
    R    &:= \set{(s(a), t(a)) | a\in Y} \\
    \sim &:= \text{reflexive, symmetric and transitive closure of $R$} \\
    C    &:= Y / \sim
  \end{alignat*}
  Note that $\class{y} := \set{ y' \in Y | (y, y') \in \sim}$ is the equivalence class of $y$ and $Y / \sim := \set{ \class{y} | y \in Y}$ is the set of all equivalence classes of $\sim$.
  Also note that no equivalence class can be empty and for $[y]$ we call $y$ a representant of this class.
  We define $i \from Y \to E$ with $i(y) := \class{y}$.
  \begin{itemize}
    \item[(a)]Show that $\sim$ is reflexive and symmetric and contains $R$.
    \item[(b)]Show that the following diagramm commutes:
      $\xymatrix{
        X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
      }$
    \item[(c)]Let
      $\xymatrix{
        X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
      }$
      be a cummutative diagramm. Show that $f$ respects equivalence, i.\,e. $\forall (y_1, y_2) \in \sim: f(y_1) = f(y_2)$.
      \begin{hint}
        ``Destruct'' $(y_1, y_2) \in \sim$ and do a $\nat$-induction.
      \end{hint}
    \item[(d)]Now define $\bar{f} \from C \to A$, such that the following diagramm commutes and show that there can only be one:
      \[ \xymatrix{
        & & A  \\
        X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[ru]^f \ar[rd]_i \\
        & & C \ar@{.>}[uu]|{\bar{f}}
      } \]
      \hint{Assume the \demph{axiom of choice.}}
    \item[(e)]Show that the fork
      $\xymatrix{
        X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
      }$
      is \emph{the} coequalizer of
      $\xymatrix{
        X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
      }$.
  \end{itemize}
\end{exercise}

\begin{answer}
  \begin{itemize}
    \item[(a)]
      This follows because for $n:=0$, $R_0$ is by construction reflexive, symmetric and cointains $R$.

    \item[(b)]
      \begin{proof}
        Let $x \in X$. We have $i(s(x)=\class{s(x)}$ and $i(t(x))=\class{t(x)}$. It remains to show that $\class{s(x)}=\class{t(x)}$. Therefore it is enough to show that $s(x) \sim t(x)$.
        This is true since $\sim$ contains $R$ and $(s(x), t(x)) \in R$ by the construction of $R$.
      \end{proof} % (b) (i)

    \item[(c)]
      \begin{proof}
        Let $(y_1, y_2) \in \sim$. Then be definition of $\sim$ we have a $n\in\nat$ such that $(y_1, y_2) \in R_n$.
        We show $\forall n\in\nat: \forall y_1,y_2 \in Y: (y_1, y_2) \in R_n \Rightarrow f (y_1) = f (y_2)$ by induction on $n\in\nat$.
        \begin{itemize}
          \item[$n=0$:]
            We have $(y_1, y_2) \in R_0$. This means $(y_1, y_2) \in R \lor (y_2, y_1) \in R \lor y_1 = y_2)$.
            We have to show $f(y_1) = f(y_2)$. Case analysis.
            \begin{itemize}
              \item[1.] $(y_1,y_2) \in R$. Then we have $y_1 = s(a) \land y_2 = t(t)$ for some $a \in X$. $f (s (a)) = f (t (a))$ holds since $f$ commutes.
              \item[2.] $(y_1,y_2) \in R$. Then we have $y_2 = s(a) \land y_1 = t(t)$ for some $a \in X$. $f (t (a)) = f (s (a))$ holds since $f$ commutes.
              \item[3.] $y_1=y_2$. Then we have $f(y_1) = f(y_2)$.
            \end{itemize}
          \item[$n \rightarrow n+1$:]
            We have $(y_1, y_2) \in R_{n+1}$. Therefore we have a $y$ such that $(y_1, y) \in R_n \land (y, y_2) \in R_n$.
            With the inductive hypositive we know $\forall y_1, y_2 \in Y: (y_1, y_2) \in R_n \Rightarrow f (y_1) = f (y_2)$.
            By instanciating the inductive hypositive we know $f (y_1) = f (y) \land f (y) = f (y_2)$.
            Thus we get $f (y_1) = f (y_2)$ \qedhere
        \end{itemize}
      \end{proof} % (b) (ii)

    \item[(d)]
      We have to assume the axiom of choice. Because every equivalence class is non-empty, we get a function
      $\choice{ \_ } \from Y/\sim \to Y$ such that $\choice{ e } \sim x$ for every equivalence class $e$ and every object $x \in e$.

      We now define $\bar{f}(e) := f (\choice e)$.

      We have to show that the diagramm commutes. Note that commutativity of this diagramm is equivalent to $\bar{f} \of i =f$.
      Let $y \in Y$. We have to show $\bar{f}(\choice{i (y)}) = f(y)$. With (c) it is enough to show $\choice{i (y)} \sim y$.
      This holds since $\choice{i (y)} = \choice{\class{y}} \sim y$, since $y \sim y$. \qedhere

      We finally have to show that $\bar{f}$ is the only function with that property.
      Let $\tilde{f} \from C \to A$ with $\tilde{f} \of i = f$. Let $c \in Y /\sim$.
      Then $\bar{f}(c) = f(\choice{c}) = \tilde{f}(i(\choice{c}) = \tilde{f}(\class{\choice{c}}) = \tilde{f}(c)$.

    \item[(e)]We know that there is only \emph{one} coequalizer (up-to isomorphism). By definition we have shown that our construction is \emph{a} coequalizer.
  \end{itemize}
\end{answer}
